\textsc{PAPERCLIP}: Associating Astronomical Observations and Natural Language with Multi-Modal Models

### 2024-02-27 paper/main.tex
Added percents to tex source
Text comments and updates

\title{\textsc{HubbleCLIP}: Associating Astronomical Observations and Natural Language with Multi-Modal Models}

% Authors must not appear in the submitted version. They should be hidden
% as long as the tmlr package is used without the [accepted] or [preprint] options.
% Non-anonymous submissions will be rejected without review.

\author{\name Siddharth Mishra-Sharma \email \href{mailto:smsharma@mit.edu}{smsharma@mit.edu} \\
      \addr The NSF AI Institute for Artificial Intelligence and Fundamental Interactions\\
      Center for Theoretical Physics, Massachusetts Institute of Technology, Cambridge, MA 02139, USA \\
      Department of Physics, Harvard University, Cambridge, MA 02138, USA
      \AND
      \name Yiding Song \email \href{mailto:ydsong@mit.edu}{ydsong@mit.edu} \\
      \addr The NSF AI Institute for Artificial Intelligence and Fundamental Interactions\\
      Department of Physics, Massachusetts Institute of Technology, Cambridge, MA 02139, USA \\
      \AND
      \name Jesse Thaler \email \href{mailto:jthaler@mit.edu}{jthaler@mit.edu} \\
      \addr The NSF AI Institute for Artificial Intelligence and Fundamental Interactions\\
      Center for Theoretical Physics, Massachusetts Institute of Technology, Cambridge, MA 02139, USA \\
}
% The \author macro works with any number of authors. Use \AND 
% to separate the names and addresses of multiple authors.

@@

\begin{abstract}
We train a multi-modal model which associates astronomical observations imaged by the \hubble Space Telescope (HST) with natural language. The model is fine-tuned from a pre-trained Contrastive Language–Image Pre-training (CLIP) model using successful proposal abstracts corresponding to HST observations, optionally summarized via guided large language model (LLM) generation. We show that the model embodies a meaningful joint representation between observations and text through experiments targeting image retrieval (i.e., retrieving the most relevant observations from a set using natural language queries) and description retrieval (i.e., querying the astrophysical object classes and science use cases most relevant to a given observation). Our study demonstrates the potential for using generalist rather than task-specific models for astrophysics research, in particular by leveraging text as an interface.
\end{abstract}

@@

Machine learning (ML) is beginning to have a significant impact in the sciences, with astrophysics being no exception. ML methods have demonstrated promise at every stage of the research pipeline, from instrument design, to data acquisition, to its analysis \citep{huertas2022dawes}. Until recently, most applications of ML within astrophysics have focused on replacing or augmenting traditional techniques with ML counterparts in order to improve performance on specific, individual tasks. The \emph{Foundation Model} paradigm, in contrast, seeks to develop generalist models which can be deployed to tackle a wide range of tasks. The paradigm has been highly successful in domains like computer vision and natural language processing, as demonstrated by the widespread adoption of tools like ChatGPT, Dall-E, Stable Diffusion, and CLIP. These models are typically pre-trained on massive amounts of unlabeled data using self-supervised learning techniques, enabling them to learn powerful representations which can be deployed on different tasks across domains. At the heart of the paradigm lies the triumph of scale -- scaling up model size, dataset size, and compute. However, foundation models can often benefit from additional training ({fine tuning}) using a relatively small amounts of domain-specific data in order to increase their usefulness when applied to specialized domains.

There is considerable interest in developing custom foundation models for the sciences \citep[e.g., ][]{batatia2023foundation,subramanian2023towards}, with astrophysics being ripe for such an effort given the availability of large amounts of publicly-available data and diverse ways of interacting with it. The multi-modality inherent to the nature of astrophysical observations, with different types of data (e.g., images, spectra, light curves, textual descriptions) often available for a given object, presents a unique opportunity. This multi-modality was recently exploited in \textsc{AstroCLIP}~\citep{lanusse2023astroclip} to obtain a joint embedding space between multi-band images and optical spectra from the Dark Energy Spectroscopic Instrument (DESI). \textsc{AstroLLaMA}~\citep{nguyen2023astrollama,perkowski2024astrollama} is another recent effort to fine-tune a publicly-available model (\textsc{Llama-2}) on astrophysics-specific textual data from the arXiv.

The Contrastive Language-Image Pre-training~\citep[CLIP;][]{radford2021learning} family of foundation models, which in their original instantiation embed images and associated captions into a common embedding space via contrastive learning, have shown strong performance and generalization capabilities on a variety of downstream tasks including zero-shot classification and image retrieval. The concept of associating diverse modalities via contrastive training has been used on other scientific domains~\citep[e.g.,][]{liu2023text,Sanchez-Fernandez2022.11.17.516915,lanusse2023astroclip,cepeda2023geoclip}, and has been shown to be effective in learning semantically meaningful joint representations.

The success of the foundation model paradigm relies in part on the ability to flexibly leverage text as a \emph{universal interface} for interacting with models. In this work, we take this outlook and train a joint representation between observations taken by the \hubble Space Telescope (HST) and natural language. We do so by leveraging associations between LLM-summarized abstracts of successful observing proposals and corresponding downstream image observations. We show that fine-tuning a pre-trained CLIP image-text model on this data results in meaningful joint representations through quantitative and qualitative evaluation.

The paper is organized as follows. In Sec.~\ref{sec:dataset}, we describe the dataset used in this work, including its curation and processing. In Sec.~\ref{sec:methodology}, we describe the methodology used to train and evaluate the model. In Sec.~\ref{sec:results}, we present the results of our experiments on image and text retrieval tasks. We discuss future prospects and conclude in Sec.~\ref{sec:conclusion}.

@@

We curate a dataset of images of \hubble observations and corresponding text descriptions. We rely on proposal abstracts from the Proposal Abstracts Catalog\footnote{\url{https://archive.stsci.edu/hst/proposal_abstracts.html}} -- a catalog of all accepted \hubble proposals to-date -- to derive captions for the observations, optionally summarizing the abstract via guided LLM generation (described in Sec.~\ref{sec:summarization} below). The HST has been operational since its launch on April 24, 1990, and we use available proposals and observations up to Cycle 30, which commenced data-taking in 2022.

Examples of images and corresponding (summarized) captions are shown in Tab.~\ref{tab:dataset}. It can be seen that these images have specific characteristics as well as artifacts particular to the nature of data-taking and processing which distinguish them from the distribution of natural images typically used for large-scale pre-training of foundation models. This further motivates the need for fine-tuning on domain-specific data.

@@

Observations corresponding to individual proposal IDs are queried through the Mikulski Archive for Space Telescopes (MAST)\footnote{\url{https://mast.stsci.edu/}} via the \package{Astroquery} \citep{2019AJ....157...98G} interface. Products of type \texttt{PREVIEW} are filtered in, corresponding to preview postcard images. We note that these are not science-grade observations, but rather lower-resolution images useful for diagnostic purposes. A maximum of 20 images are downloaded per proposal ID, selected at random, in order to avoid biasing the model towards proposals with a larger number of observations and survey-style campaigns. Images are centered and resized to a resolution-per-side of 512 pixels before. Color previews (i.e., observations taken with multiple wavelength filters assigned to individual RGB channels) are manually excluded via a filename filter in order to maintain consistency across the dataset; models trained on datasets with color images included were observed to show worse performance on generalization metrics. If no appropriate images corresponding to an abstract are found, it is excluded from the dataset.

In total, 31,859 images corresponding to 4,438 abstracts are included in the fine-tuning dataset. 3,194 images are held out for validation, with no abstract being common between training and validation sets in order to ensure an independent set of text-image pairs for evaluation.

We note that some fraction of the image-caption pairs in the constructed dataset will center instrumentation and/or calibration rather than primarily scientific content. We choose to not filter out these pairs from our dataset, in order to have a larger sample of HST observations that the image encoder can leverage to adapting to the distinctive characteristics of \hubble images.

@@

Raw proposal abstracts summarize the corresponding successful HST observing proposals, which intend to make the case for allocating \hubble telescope time towards a particular set of observations. These abstracts are written in a diversity of styles, formats, and lengths, being highly variable in the nature of content as well. Although the abstracts can be used as-is as image captions, we experiment with summarizing them via guided large language model (LLM) generation to standardize the captions used for fine-tuning the CLIP model. The goal is to summarize the objects and phenomena, as well as potential downstream science use cases corresponding to the HST observations in order to increase the strength of the association signal between text and images.

The method from \cite{willard2023efficient} is used to produce an LLM-generated summary of the abstract conforming to a particular schema, specified in JSON format. The schema is designed to represent a list of the objects (e.g., `Type Ia supernova') and phenomena (e.g., `gravitational lensing'), as well as potential downstream science uses cases (e.g., `set constraints on supernova explosion models') that could correspond to the eventual imaged observation given the abstract text, with a minimum of 1 and a maximum of 5 elements per list.

The procedure guides the generation of LLM outputs while ensuring that the schema is respected at every step in the autoregressive generation process by masking out tokens that would violate the intended format. By framing the problem in terms of transitions between a set of finite-state machines, \cite{willard2023efficient} showed that guided generation can be performed with negligible overhead compared to unconstrained generation. This ensures that the output of the LLM strictly conforms to the format of the following example:
\begin{lstlisting}[language=Python]
{
  'objects_and_phenomena': ['star forming galaxy', 'lensed galaxy', ...], 
  'science_use_cases': ['measure lensing magnification', 'probe spectral energy distributions', ...]
}
\end{lstlisting}
which is then used to construct the summarized caption by combining the two key elements. Examples of raw abstract snippets and corresponding LLM-generated summaries are shown in Tab.~\ref{tab:datasetsumm}. See App.~\ref{app:guided-generation} for a more detailed description of the guidance generation process.

We use the open-weights, instruction-tuned model \textsc{Mixtral-8x7B-Instruct}~\citep{jiang2024mixtral} to generate the summaries, with guided generation performed using the \package{Outlines}\footnote{\url{https://github.com/outlines-dev/outlines}} package. Further details on the summarization procedure, including the prompts and schema used, are provided in App.~\ref{app:summarization}.

We emphasize that the goal of summarization-via-guided-generation is to increase the signal between text and images by standardizing the captions used for fine-tuning the CLIP model, and compare the quantitative performance of the model vs using the raw abstracts in Sec.~\ref{sec:results}. We also note that, even after summarization, the association signal is expected to be noisy, since aspects of the summarized caption may not be directly descriptive of the observed images, as can be seen from Tab.~\ref{tab:dataset}.

Finally, in order to test a further compression of the proposal abstracts, we use LLM-guided generation to also produce a list of single-concept summaries (e.g., `irregular galaxy', `quasars', `Galactic bulge', \ldots), and test whether this can lead to meaningful, generalizable learned associations with observations. The prompts and schemata for generating these discrete categories and assigning observations to them are described in Apps.~\ref{app:singleconcept} and \ref{app:singleconceptassignments} respectively.

@@

\begin{table}[h!]
      \centering
      \begin{tabular}{m{0.21\textwidth} p{1.9cm} p{1.9cm} m{8cm}}
          \toprule
          \centering \bfseries Image & \centering \bfseries Obs. cycle & \centering \bfseries Prop. ID & \centering \bfseries LLM-extracted summary \tabularnewline
          \midrule
          \centering \includegraphics[width=0.2\textwidth]{\thedatafolder/img_0.pdf} & \centering \input{\thedatafolder/cycle_0.txt} & \centering \input{\thedatafolder/id_0.txt} &  {\scriptsize \input{\thedatafolder/sum_0.txt}} \tabularnewline
          \midrule
          \centering \includegraphics[width=0.2\textwidth]{\thedatafolder/img_1.pdf} & \centering \input{\thedatafolder/cycle_1.txt} & \centering \input{\thedatafolder/id_1.txt} &  {\scriptsize \input{\thedatafolder/sum_1.txt}} \tabularnewline
          \midrule
          \centering \includegraphics[width=0.2\textwidth]{\thedatafolder/img_2.pdf} & \centering \input{\thedatafolder/cycle_2.txt} & \centering \input{\thedatafolder/id_2.txt} &  {\scriptsize \input{\thedatafolder/sum_2.txt}} \tabularnewline
          \midrule
          \centering \includegraphics[width=0.2\textwidth]{\thedatafolder/img_3.pdf} & \centering \input{\thedatafolder/cycle_3.txt} & \centering \input{\thedatafolder/id_3.txt} &  {\scriptsize \input{\thedatafolder/sum_3.txt}} \tabularnewline
          \bottomrule
      \end{tabular}
      \caption{Examples of images and corresponding captions, constructed using the LLM-extracted summaries. The CLIP model is fine-tuned using these text-image associations.}
      \label{tab:dataset}
  \end{table}

\begin{landscape}
      \begin{table}[h!]
          \centering
          \begin{tabular}{m{1.8cm} m{8cm} m{5cm} m{6cm}}
              \toprule
              \bfseries Prop. ID & \centering\arraybackslash \bfseries Proposal abstract & \multicolumn{2}{c}{\bfseries LLM-extracted summary} \tabularnewline
              \cmidrule(r){3-4}
              & & \centering\arraybackslash \bfseries Objects and phenomena & \centering\arraybackslash \bfseries Science use cases \tabularnewline
              \midrule
              \input{\thedatafolder/id1_0.txt} & {\scriptsize \input{\thedatafolder/abs1_0.txt}} & {\scriptsize \input{\thedatafolder/obj1_0.txt}} & {\scriptsize \input{\thedatafolder/sci1_0.txt}} \tabularnewline
              \midrule
              \input{\thedatafolder/id1_1.txt} & {\scriptsize \input{\thedatafolder/abs1_1.txt}} & {\scriptsize \input{\thedatafolder/obj1_1.txt}} & {\scriptsize \input{\thedatafolder/sci1_1.txt}} \tabularnewline
              \midrule
              \input{\thedatafolder/id1_2.txt} & {\scriptsize \input{\thedatafolder/abs1_2.txt}} & {\scriptsize \input{\thedatafolder/obj1_2.txt}} & {\scriptsize \input{\thedatafolder/sci1_2.txt}} \tabularnewline
              \midrule
              \input{\thedatafolder/id1_3.txt} & {\scriptsize \input{\thedatafolder/abs1_3.txt}} & {\scriptsize \input{\thedatafolder/obj1_3.txt}} & {\scriptsize \input{\thedatafolder/sci1_3.txt}} \tabularnewline
              \bottomrule
          \end{tabular}
          \caption{Examples of the initial parts of raw proposal abstracts (second column) and LLM (\textsc{Mixtral-8x7B})-extracted summaries (rightmost two columns), separately extracting objects and phenomena as well as potential downstream science use cases.}
          \label{tab:datasetsumm}
      \end{table}
  \end{landscape}

@@

Our goal is to learn a semantically  meaningful joint representation between HST observation images and natural language. We leverage the strong generalization capabilities demonstrated by pre-trained CLIP models and adapt these to work with domain-specific \hubble data via fine tuning.

@@

CLIP \citep[Contrastive Language-Image Pretraining;][]{radford2021learning} is a multi-modal model pre-trained on a large corpus of image-text pairs via weak supervision using a contrastive loss. Given a minibatch $\mathcal{B}$ of $|\mathcal{B}|$ image-text pairs $\{(I_i, T_i)\}$, the goal is to align the learned representations of corresponding (positive) pairs $(I_i, T_i)$ while repelling the representations of unaligned (negative) pairs $(I_i, T_{j\neq i})$. Image and text encoders $f: I \rightarrow \mathbb R^{n_\text{emb}}$ and $g: T \rightarrow \mathbb R^{n_\text{emb}}$ are used to map images and text to a common embedding space of dimension $n_\text{emb}$. The standard softmax-based bidirectional variation of the InfoNCE~\citep{oord2018representation} contrastive loss function was introduce for training CLIP-style architectures \citep{radford2021learning}
\begin{equation}
  \mathcal{L}(\mathcal{B})=-\frac{1}{2|\mathcal{B}|} \sum_{i=1}^{|\mathcal{B}|}\left(\log \frac{e^{x_i \cdot y_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{x_i \cdot y_j / \tau}}+\log \frac{e^{x_i \cdot y_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{x_j \cdot y_i / \tau}}\right)\label{eq:softmax_loss}
\end{equation}
where ${x}_i={f\left(I_i\right)}/{\left\|f\left(I_i\right)\right\|}$ and ${y}_i={g\left(T_i\right)}/{\left\|g\left(T_i\right)\right\|}$ are the normalized representations of the $i$-th image and text, respectively, and $\tau$ is a learnable temperature hyperparameter. Note that this loss treats the image and text representations symmetrically, ensuring that the two modalities are on the same footing.

We use the CLIP-ViT-B/16 \citep{radford2021learning} variant as the base pre-trained CLIP model. This model uses a 12-layer, 12-head, 768-embedding dimension vision transformer as the image encoder and a 12-layer, 8-head, 512-embedding dimension sequence transformer as the text backbone. The text encoder has a maximum length of 77 tokens and the image encoder a native resolution of $224\times224$. Linear projection layers map the outputs of the image and text encoders to a common embedding space of dimension $n_\text{emb}=512$. In total, the model has 149,620,737 trainable parameters. The model was originally trained on 400 million image-text pairs from internet data.

@@

Figure~\ref{fig:retrieval_acc} shows the contrastive loss (left) and the top-10\% retrieval accuracy on the held out validation set over the course of training, for different training configurations considered. The purple lines show the metrics evaluated when training with batches where the image-text associations are randomly shuffled, serving as a baseline. This baseline is seen to do on par with random expectation, unlike the others, validating the presence of a significant association signal between images and text in the dataset.
%
Interestingly, the base pre-trained model performs better than random expectation, with a top-10\% retrieval accuracy of $\sim 15\%$. We therefore compare the qualitative performance of the base model with the fine-tuned model on downstream retrieval tasks.

>>>>

\title{\textsc{PAPERCLIP}: Associating Astronomical Observations and Natural Language with Multi-Modal Models}

\author{\name Siddharth Mishra-Sharma \email \href{mailto:smsharma@mit.edu}{smsharma@mit.edu} \\
      \addr The NSF AI Institute for Artificial Intelligence and Fundamental Interactions\\
      Center for Theoretical Physics, Massachusetts Institute of Technology, Cambridge, MA 02139, USA \\
      Department of Physics, Harvard University, Cambridge, MA 02138, USA
      \AND
      \name Yiding Song \email \href{mailto:ydsong@mit.edu}{ydsong@mit.edu} \\
      \addr The NSF AI Institute for Artificial Intelligence and Fundamental Interactions\\
      Department of Physics, Massachusetts Institute of Technology, Cambridge, MA 02139, USA \\
      \AND
      \name Jesse Thaler \email \href{mailto:jthaler@mit.edu}{jthaler@mit.edu} \\
      \addr The NSF AI Institute for Artificial Intelligence and Fundamental Interactions\\
      Center for Theoretical Physics, Massachusetts Institute of Technology, Cambridge, MA 02139, USA \\
}

@@

\begin{abstract}
We present PAPERCLIP (Proposal Abstracts Provide an Effective Representation for Contrastive Language-Image Pretraining), a method which associates astronomical observations imaged by astronomical surveys like the \hubble Space Telescope (HST) with natural language. The model is fine-tuned from a pre-trained Contrastive Language–Image Pre-training (CLIP) model using successful proposal abstracts corresponding to HST observations, optionally summarized via guided generation using large language models (LLMs). We show that the model embodies a meaningful joint representation between observations and text through experiments targeting image retrieval (i.e., retrieving the most relevant observations from a set using natural language queries) and description retrieval (i.e., querying the astrophysical object classes and science use cases most relevant to a given observation). Our study demonstrates the potential for using generalist rather than task-specific models for astrophysics research, in particular by leveraging text as an interface.
\end{abstract}

@@

\SM{Add TOC.}
\SM{Change PAPERCLIP everywhere.}
\SM{Add some qualitative evals with the abstract model.}

Machine learning (ML) is beginning to have a significant impact in the sciences, with astrophysics being no exception.
%
ML methods have demonstrated promise at every stage of the research pipeline, from instrument design, to data acquisition, to its analysis \citep{huertas2022dawes}.
%
Until recently, most applications of ML within astrophysics have focused on replacing or augmenting traditional techniques with ML counterparts in order to improve performance on specific, individual tasks.
%
The \emph{Foundation Model} paradigm, in contrast, seeks to develop generalist models which can be deployed to tackle a wide range of tasks.
%
The paradigm has been highly successful in domains like computer vision and natural language processing, as demonstrated by the widespread adoption of tools like ChatGPT, Dall-E, Stable Diffusion, and CLIP.
%
These models are typically pre-trained on massive amounts of unlabeled data using self-supervised learning techniques, enabling them to learn powerful representations which can be deployed on different tasks across domains.
%
At the heart of the paradigm lies the triumph of scale -- scaling up model size, dataset size, and compute.
%
However, foundation models can often benefit from additional training ({fine tuning}) using a relatively small amounts of domain-specific data in order to increase their usefulness when applied to specialized domains.

There is considerable interest in developing custom foundation models for the sciences \citep[e.g., ][]{batatia2023foundation,subramanian2023towards}, with astrophysics being ripe for such an effort given the availability of large amounts of publicly-available data and diverse ways of interacting with it.
%
The multi-modality inherent to the nature of astrophysical observations, with different types of data (e.g., images, spectra, light curves, textual descriptions) often available for a given object, presents a unique opportunity.
%
This multi-modality was recently exploited in \textsc{AstroCLIP}~\citep{lanusse2023astroclip} to obtain a joint embedding space between multi-band images and optical spectra from the Dark Energy Spectroscopic Instrument (DESI).
%
\textsc{AstroLLaMA}~\citep{nguyen2023astrollama,perkowski2024astrollama} is another recent effort to fine-tune a publicly-available model (\textsc{Llama-2}) on astrophysics-specific textual data from the arXiv.

The Contrastive Language-Image Pre-training~\citep[CLIP;][]{radford2021learning} family of foundation models, which in their original instantiation embed images and associated captions into a common embedding space via contrastive learning, have shown strong performance and generalization capabilities on a variety of downstream tasks including zero-shot classification and image retrieval.
%
The concept of associating diverse modalities via contrastive training has been used on other scientific domains~\citep[e.g.,][]{liu2023text,Sanchez-Fernandez2022.11.17.516915,lanusse2023astroclip,cepeda2023geoclip}, and has been shown to be effective in learning semantically meaningful joint representations.

\SM{Put in this work we at the start of the paragraph. Also put the paragraph right after AstroLlama, reworking some of the logic.} \SM{Define PAPERCLIP, funny footnote about PAPERCLIFT.}

The success of the foundation model paradigm relies in part on the ability to flexibly leverage text as a \emph{universal interface} for interacting with models.
%
In this work, we take this outlook and train a joint representation between observations taken by the \hubble Space Telescope (HST) and natural language.
%
We do so by leveraging associations between LLM-summarized abstracts of successful observing proposals and corresponding downstream image observations.
%
We show that fine-tuning a pre-trained CLIP image-text model on this data results in meaningful joint representations through quantitative and qualitative evaluation.

The paper is organized as follows.
%
In Sec.~\ref{sec:dataset}, we describe the dataset used in this work, including its curation and processing.
%
In Sec.~\ref{sec:methodology}, we describe the methodology used to train and evaluate the model.
%
In Sec.~\ref{sec:results}, we present the results of our experiments on image and text retrieval tasks.
%
We discuss future prospects and conclude in Sec.~\ref{sec:conclusion}.

@@

We curate a dataset of images of \hubble observations and corresponding text descriptions.
%
We rely on proposal abstracts from the Proposal Abstracts Catalog\footnote{\url{https://archive.stsci.edu/hst/proposal_abstracts.html}} -- a catalog of all accepted \hubble proposals to-date -- to derive captions for the observations, optionally summarizing the abstract via guided LLM generation (described in Sec.~\ref{sec:summarization} below).
%
The HST has been operational since its launch on April 24, 1990, and we use available proposals and observations up to Cycle 30, which commenced data-taking in 2022.
%

Examples of images and corresponding (summarized) captions are shown in Tab.~\ref{tab:dataset}. \SM{Push table earlier.}
%
It can be seen that these images have specific characteristics as well as artifacts particular to the nature of data-taking and processing which distinguish them from the distribution of natural images typically used for large-scale pre-training of foundation models.
%
This further motivates the need for fine-tuning on domain-specific data.

@@

Observations corresponding to individual proposal IDs are queried through the Mikulski Archive for Space Telescopes (MAST)\footnote{\url{https://mast.stsci.edu/}} via the \package{Astroquery} \citep{2019AJ....157...98G} interface.
%
Products of type \texttt{PREVIEW} are filtered in, corresponding to preview postcard images.
%
We note that these are not science-grade observations, but rather lower-resolution images useful for diagnostic purposes.
%
A maximum of 20 images are downloaded per proposal ID, selected at random, in order to avoid biasing the model towards proposals with a larger number of observations and survey-style campaigns.
%
Images are centered and resized to a resolution-per-side of 512 pixels.
%
Color previews (i.e., observations taken with multiple wavelength filters assigned to individual RGB channels) are manually excluded via a filename filter in order to maintain consistency across the dataset; models trained on datasets with color images included were observed to show worse performance on generalization metrics.
%
If no appropriate images corresponding to an abstract are found, it is excluded from the dataset.

In total, 31,859 images corresponding to 4,438 abstracts are included in the fine-tuning dataset.
%
3,194 images are held out for validation, with no abstract being common between training and validation sets in order to ensure an independent set of text-image pairs for evaluation. \SM{Put number of held out abstracts here}

We note that some fraction of the image-caption pairs in the constructed dataset will primarily concern instrumentation and/or calibration rather than scientific content.
%
We choose to not filter out these pairs from our dataset, in order to have a larger sample of HST observations that the image encoder can leverage to adapting to the distinctive characteristics of \hubble images.

@@

Raw proposal abstracts summarize the corresponding successful HST observing proposals, which intend to make the case for allocating \hubble telescope time towards a particular set of observations.
%
These abstracts are written in a diversity of styles, formats, and lengths, being highly variable in the nature of content as well.
%
Although the abstracts can be used as-is as image captions, we experiment with summarizing them via guided large language model (LLM) generation to standardize the captions used for fine-tuning the CLIP model.
%
The goal is to summarize the objects and phenomena, as well as potential downstream science use cases corresponding to the HST observations in order to increase the strength of the association signal between text and images.

The method from \cite{willard2023efficient} is used to produce an LLM-generated summary of the abstract conforming to a particular schema, specified in JSON format.
%
The schema is designed to represent a list of the objects (e.g., `Type Ia supernova') and phenomena (e.g., `gravitational lensing'), as well as potential downstream science uses cases (e.g., `set constraints on supernova explosion models') that could correspond to the eventual imaged observation given the abstract text, with a minimum of 1 and a maximum of 5 elements per list. \SM{Emphasize that the schema ensures the format, but the prompt is necessary for it to make sense.}

The procedure guides the generation of LLM outputs while ensuring that the schema is respected at every step in the autoregressive generation process by masking out tokens that would violate the intended format.
%
By framing the problem in terms of transitions between a set of finite-state machines, \cite{willard2023efficient} showed that guided generation can be performed with negligible overhead compared to unconstrained generation.
%
This ensures that the output of the LLM strictly conforms to the format of the following example:
\begin{lstlisting}[language=Python]
{
  'objects_and_phenomena': ['star forming galaxy', 'lensed galaxy', ...], 
  'science_use_cases': ['measure lensing magnification', 'probe spectral energy distributions', ...]
}
\end{lstlisting}
which is then used to construct the summarized caption by combining the two key elements.
%
Examples of raw abstract snippets and corresponding LLM-generated summaries are shown in Tab.~\ref{tab:datasetsumm}.
%
See App.~\ref{app:guided-generation} for a more detailed description of the guidance generation process.

We use the open-weights, instruction-tuned model \textsc{Mixtral-8x7B-Instruct}~\citep{jiang2024mixtral} to generate the summaries, with guided generation performed using the \package{Outlines}\footnote{\url{https://github.com/outlines-dev/outlines}} package.
%
Further details on the summarization procedure, including the prompts and schema used, are provided in App.~\ref{app:summarization}.

We emphasize that the goal of summarization-via-guided-generation is to increase the signal between text and images by standardizing the captions used for fine-tuning the CLIP model, and compare the quantitative performance of the model vs using the raw abstracts in Sec.~\ref{sec:results}.
%
We also note that, even after summarization, the association signal is expected to be noisy, since aspects of the summarized caption may not be directly descriptive of the observed images, as can be seen from Tab.~\ref{tab:dataset}.

Finally, in order to test a further compression of the proposal abstracts, we use LLM-guided generation to also produce a list of single-concept summaries (e.g., `irregular galaxy', `quasars', `Galactic bulge', \ldots), and test whether this can lead to meaningful, generalizable learned associations with observations.
%
The prompts and schemata for generating these discrete categories and assigning observations to them are described in Apps.~\ref{app:singleconcept} and \ref{app:singleconceptassignments} respectively. \SM{Each abstract is assigned one summary. The generation of the single concepts is different }

@@

\begin{table}[h!]
      \centering
      \begin{tabular}{m{0.21\textwidth} p{1.9cm} p{1.9cm} m{8cm}}
          \toprule
          \centering \bfseries Image & \centering \bfseries Obs. cycle & \centering \bfseries Prop. ID & \centering \bfseries LLM-extracted summary \tabularnewline
          \midrule
          \centering \includegraphics[width=0.2\textwidth]{\thedatafolder/img_0.pdf} & \centering \input{\thedatafolder/cycle_0.txt} & \centering \input{\thedatafolder/id_0.txt} &  {\scriptsize \input{\thedatafolder/sum_0.txt}} \tabularnewline
          \midrule
          \centering \includegraphics[width=0.2\textwidth]{\thedatafolder/img_1.pdf} & \centering \input{\thedatafolder/cycle_1.txt} & \centering \input{\thedatafolder/id_1.txt} &  {\scriptsize \input{\thedatafolder/sum_1.txt}} \tabularnewline
          \midrule
          \centering \includegraphics[width=0.2\textwidth]{\thedatafolder/img_2.pdf} & \centering \input{\thedatafolder/cycle_2.txt} & \centering \input{\thedatafolder/id_2.txt} &  {\scriptsize \input{\thedatafolder/sum_2.txt}} \tabularnewline
          \midrule
          \centering \includegraphics[width=0.2\textwidth]{\thedatafolder/img_3.pdf} & \centering \input{\thedatafolder/cycle_3.txt} & \centering \input{\thedatafolder/id_3.txt} &  {\scriptsize \input{\thedatafolder/sum_3.txt}} \tabularnewline
          \bottomrule
      \end{tabular}
      \caption{Examples of images and corresponding captions, constructed using the LLM-extracted summaries.
%
The CLIP model is fine-tuned using these text-image associations. \SM{Semicolon separates, but the model doesn't know about it.}}
      \label{tab:dataset}
  \end{table}

\begin{landscape}
      \begin{table}[h!]
          \centering
          \begin{tabular}{m{1.8cm} m{8cm} m{5cm} m{6cm}}
              \toprule
              \bfseries Prop. ID & \centering\arraybackslash \bfseries Proposal abstract & \multicolumn{2}{c}{\bfseries LLM-extracted summary} \tabularnewline
              \cmidrule(r){3-4}
              & & \centering\arraybackslash \bfseries Objects and phenomena & \centering\arraybackslash \bfseries Science use cases \tabularnewline
              \midrule
              \input{\thedatafolder/id1_0.txt} & {\scriptsize \input{\thedatafolder/abs1_0.txt}} & {\scriptsize \input{\thedatafolder/obj1_0.txt}} & {\scriptsize \input{\thedatafolder/sci1_0.txt}} \tabularnewline
              \midrule
              \input{\thedatafolder/id1_1.txt} & {\scriptsize \input{\thedatafolder/abs1_1.txt}} & {\scriptsize \input{\thedatafolder/obj1_1.txt}} & {\scriptsize \input{\thedatafolder/sci1_1.txt}} \tabularnewline
              \midrule
              \input{\thedatafolder/id1_2.txt} & {\scriptsize \input{\thedatafolder/abs1_2.txt}} & {\scriptsize \input{\thedatafolder/obj1_2.txt}} & {\scriptsize \input{\thedatafolder/sci1_2.txt}} \tabularnewline
              \midrule
              \input{\thedatafolder/id1_3.txt} & {\scriptsize \input{\thedatafolder/abs1_3.txt}} & {\scriptsize \input{\thedatafolder/obj1_3.txt}} & {\scriptsize \input{\thedatafolder/sci1_3.txt}} \tabularnewline
              \bottomrule
          \end{tabular}
          \caption{Examples of the initial parts of raw proposal abstracts (second column) and LLM (\textsc{Mixtral-8x7B})-extracted summaries (rightmost two columns), separately extracting objects and phenomena as well as potential downstream science use cases. \SM{Cite all proposal.} \SM{Add a year with the obs. cycle.}.}
          \label{tab:datasetsumm}
      \end{table}
  \end{landscape}

@@

Our goal is to learn a semantically  meaningful joint representation between HST observation images and natural language.
%
We leverage the strong generalization capabilities demonstrated by pre-trained CLIP models and adapt these to work with domain-specific \hubble data via fine tuning.

@@

CLIP \citep[Contrastive Language-Image Pretraining;][]{radford2021learning} is a multi-modal model pre-trained on a large corpus of image-text pairs via weak supervision using a contrastive loss.
%
Given a minibatch $\mathcal{B}$ of $|\mathcal{B}|$ image-text pairs $\{(I_i, T_i)\}$, the goal is to align the learned representations of corresponding (positive) pairs $(I_i, T_i)$ while repelling the representations of unaligned (negative) pairs $(I_i, T_{j\neq i})$.
%
Image and text encoders $f: I \rightarrow \mathbb R^{n_\text{emb}}$ and $g: T \rightarrow \mathbb R^{n_\text{emb}}$ are used to map images and text to a common embedding space of dimension $n_\text{emb}$.
%
The standard softmax-based bidirectional variation of the InfoNCE~\citep{oord2018representation} contrastive loss function was introduce for training CLIP-style architectures \citep{radford2021learning}
\begin{equation}
  \mathcal{L}(\mathcal{B})=-\frac{1}{2|\mathcal{B}|} \sum_{i=1}^{|\mathcal{B}|}\left(\log \frac{e^{x_i \cdot y_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{x_i \cdot y_j / \tau}}+\log \frac{e^{x_i \cdot y_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{x_j \cdot y_i / \tau}}\right)\label{eq:softmax_loss}
\end{equation}
where ${x}_i={f\left(I_i\right)}/{\left\|f\left(I_i\right)\right\|}$ and ${y}_i={g\left(T_i\right)}/{\left\|g\left(T_i\right)\right\|}$ are the normalized representations of the $i$-th image and text, respectively, and $\tau$ is a learnable temperature hyperparameter.
%
Note that this loss treats the image and text representations symmetrically, ensuring that the two modalities are on the same footing.

We use the CLIP-ViT-B/16 \citep{radford2021learning} variant as the base pre-trained CLIP model.
%
This model uses a 12-layer, 12-head, 768-embedding dimension vision transformer as the image encoder and a 12-layer, 8-head, 512-embedding dimension sequence transformer as the text backbone.
%
The text encoder has a maximum length of 77 tokens and the image encoder a native resolution of $224\times224$.
%
Linear projection layers map the outputs of the image and text encoders to a common embedding space of dimension $n_\text{emb}=512$.
%
In total, the model has 149,620,737 trainable parameters.
%
The model was originally trained on 400 million image-text pairs from internet data. \SM{$\tau$ was initialized to the pre-trained value.} \SM{HuggingFace link?} \SM{What area is maintained.} \SM{It's only 90 deg rotations.}
%

@@

Figure~\ref{fig:retrieval_acc} shows the contrastive loss (left) and the top-10\% retrieval accuracy (right) on the held out validation set over the course of training, for different training configurations considered.
%
The purple lines show the metrics evaluated when training with batches where the image-text associations are randomly shuffled, serving as a baseline.
%
This baseline is seen to do on par with random expectation, unlike the others, validating the presence of a significant association signal between images and text in the dataset.
%
Interestingly, the base pre-trained model performs better than random expectation, with a top-10\% retrieval accuracy of $\sim 15\%$.
%
We therefore compare the qualitative performance of the base model with the fine-tuned model on downstream retrieval tasks.
